{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-22 thesis/quicksand.tex
data-race

This chapter presents Quicksand,
an execution framework for model checking to manage this trade-off at run-time.
Given a fixed CPU budget,
representing the user's patience for testing,
Quicksand dynamically alters its preemption strategy based on data race analysis
(\cite{tsan,fasttrack}, \cref{sec:landslide-datarace})
and optimizes the size of state spaces on the fly,
guided by state space estimation
(\cite{estimation}, \cref{sec:landslide-estimate}),
to best match that budget.
I will discuss the trade-off inherent in number of preemption points used
(\cref{sec:quicksand-motivation}),
introduce {\em Iterative Deepening},
the algorithm that Quicksand uses to automatically navigate that trade-off
(\cref{sec:quicksand-id}),
prove its soundness relative to the more expensive full verification approach
(\cref{sec:quicksand-soundness}),
and present a large evaluation of Quicksand against several
state-of-the-art approaches % implemented under Landslide
in which Quicksand performs best on both bug-finding and verification
(\cref{sec:quicksand-eval}).
% justifying the claim that model checkers should navigate said tradeoff at runtime
}

@@

\qrevisionminor{How should a model checker know to instrument this particular data race for preemption
in order to find the assertion failure lurking underneath?}
Committing in advance to
preempt on every instruction is certain to include this data race,
but invites massive state space explosion.
Even as DPOR helps to skip equivalent interleavings of non-conflicting transitions,
DPOR itself is $O(n^2)$ in the number of preemption points in a single execution,
which is not compatible with such an approach.
Accordingly, stateless model checkers must find more efficient ways to be able to uncover bugs such as these.
%
In related work,
Portend~\cite{portend} proposed to combine data race analysis with preemption-driven artificial scheduling,
although it obtains its data-race candidates from a stand-alone, single-pass analysis.
In order to
%make sure we
identify every
%one of a program's
data race that could possibly arise under the given test case,
a model checker must check many different interleavings to begin with,
perform the Portend approach for every data race it finds,
which may in turn uncover more data races
(hidden in flow control paths reachable only through interleavings of the first race, perhaps),
and then continue model checking those multiple races together
in a
%sort of
bidirectional feedback loop between the two algorithms.
Upcoming,
\cref{sec:quicksand-id} will show how I achieve this in Quicksand, and
\cref{sec:quicksand-soundness} will justify the technique's formal verification power.
In the evaluation, %(\cref{sec:quicksand-eval}),
\qrevisionminor{\cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}
will show that Quicksand strikes a healthy balance between fast bug-finding and full verification,
and
\cref{sec:quicksand-eval-nondets}
%in particular
will justify the need
for such a feedback loop %between model checking and data race analysis.
by showing that many
data races require model checking to reliably detect.
%bugs require nondeterministic data races to expose.
}

@@

For the remainder of the chapter, I will use Iterative Deepening to refer to the algorithm in the abstract,
which could in principle apply to any stateless model checking domain,
and Quicksand to refer to the specific implementation,
which relies on data race analysis and specific heuristics to optimize its testing approach
for kernels and thread libraries.
I will also henceforth refer to each unique set of preemption points as a {\em job}.

@@

%Note that
Iterative Deepening is a wrapper algorithm around stateless model checking.
A model checker is still used to test each state space, and other reduction techniques such as DPOR
are still applicable in each.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
This chapter
will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
The next section (\cref{sec:quicksand-soundness})
will prove that in fact,
in addition to statically-known synchronization preemption points,
this suffices to provide at least as strong verification guarantees as any other possible preemption point set.

@@

However, in many tests of even modestly-sized programs,
full verification is not feasible,
and focusing on the maximal state space alone is likely to be fruitless.
%
Hence, Iterative Deepening also allows for prioritizing subset jobs
based on number of preemption points, ETA, and whether data race candidates are included among their preemption points.
It relies on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing a large fraction of interleavings for each.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only her total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.
Quicksand seeks to maximize completed state spaces,
as each one serves as a guarantee that all possible interleavings therein were tested;
\cref{sec:quicksand-discussion-partial} discusses some limitations of this approach.
The next three subsections will show how to schedule these smaller jobs
based on their preemption points and ETAs.

@@

For kernel-level testing, interrupt-disabling is analogous to locking,
so preemptions must also be introduced
just before a disable-interrupt opcode (on x86, {\tt cli})
and just after interrupts are re-enabled (on x86, {\tt sti}).
During data race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
(note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
\footnote{Some kernels disable preemption without disabling interrupts,
which can be communicated to the model checker using manual annotations,
and must be treated similarly.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}
Quicksand is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, it doesn't test every intermediate subset such as $\{lock,sti\}$,
which would result in $2^p$ jobs right off the bat,
although this could potentially be improved in future work (\cref{sec:quicksand-discussion-subsets}).

@@

As discussed in \cref{sec:quicksand-pps},
%runtime data race detection (\cref{sec:background-datarace})
%may find candidate unsynchronized memory conflicts that should be investigated further.
data races may beget new interleavings not reachable by preempting on synchronization API boundaries alone.
Because each data race indicates an access pair that can interleave at instruction granularity,
different program behaviour may arise if the threads are preempted just before the racing instructions,
some of which the programmer may not have even expected, i.e., be bugs,
and it is logical to apply model checking to find or verify the absence of such bugs.
%it is logical to re-execute the test and issue preemptions just before those instructions
%to test alternate thread interleavings.

@@

\begin{figure}[t]
	\begin{center}
        \includegraphics[width=0.9\textwidth]{dr-jobs-v2.pdf}
	\end{center}
	\caption[Quicksand incorporates data races as new preemption points at run-time.]
		{Quicksand incorporates data races as new preemption points at run-time
		by managing the exploration of multiple state spaces,
		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
                %When an access pair is reported as a data race candidate,
                When a data race is reported,
                a new preemption point is added for each involved memory access,
                and new jobs are added for later testing,
                corresponding to different combinations thereof with the existing preemption points.}
        \label{fig:new-dr-jobs}
\end{figure}

@@

% TODO: fix syntax hilight
% TODO check figure placement
% TODO check figure camption
\begin{figure}[t]
	\begin{center}
	\begin{tabular}{rll}
		& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
		& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
		\\
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		2 & \texttt{\hilight{olivegreen}{free}(x);} \\
		3 & & \texttt{\hilight{commentblue}{// x's memory reallocated}} \\
		4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
		6 & & \texttt{publish(y);} \\
		7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
	\end{tabular}
	\end{center}
	% TODO: short ver
	\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
	\label{fig:recycle}
\end{figure}

@@

The ability to disregard reallocation false positives is unique to Iterative Deepening.
When limited to a single test execution, suppressing any data race candidate matching this pattern is unsound.
Consider the more unusual program in Figure~\ref{fig:recycle-bug},
in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
Here, reordering lines~6-7 before line~3 will allow {\tt x} and {\tt x2} to race.
Discarding the data race report as a false positive after checking just this one execution
would overlook such a bug,
but Iterative Deepening is guaranteed to explore the alternate interleaving,
in which the true data race will show up without {\tt free()} and {\tt malloc()} interposing,
so it is safe to suppress at first, as I will prove in \cref{sec:quicksand-realloc}.
Moreover,
in the context of Iterative Deepening, being able to discard certain data race candidates
allows Quicksand to skip exploring some entire state spaces,
and hence run fewer Landslides overall;
this is analogous to DPOR's ability to skip equivalent interleavings within a single Landslide instance.
Upcoming in the evaluation, \cref{sec:quicksand-eval-verif}'s \Cref{tab:drstatistix}
will show how many redundant state spaces Quicksand is able to prune with this technique.

@@

% TODO: check spcaing between defs/lemmas/etc and surrounding text
\begin{definition}[Reachability]
	A data race candidate, and its associated preemption point(s),
	are reachable if it will be identified by a model checker
	configured to preempt only on already-reachable preemption points.
\end{definition}

@@

If the verbose option ({\tt -v}) is supplied,
Quicksand will also print one line per interleaving tested by all its Landslide instances,
showing the number of branches tested so far, the estimated percent progress, and the ETA,
as shown earlier in \cref{sec:landslide-estimate}.
This produces a lot more output, and can make progress reports hard to read as they scroll off the screen quickly,
but the author personally finds this mode less disorienting than ten seconds of pure silence between each progress report.
Of course, future work could improve this with a GUI, or at least a split-screen console view.
It will also cause the progress reports to report more detailed information,
such as which preemption points are nondeterministic data-races (\cref{sec:quicksand-pps})
and number of reallocation false positives suppressed (\cref{sec:quicksand-id-realloc}).

@@

The other end of the messaging protocol (\cref{sec:quicksand-impl-mc})
is implemented in {\tt messaging.c}.
When Quicksand suspends Landslide,
it detects how much time it spent asleep,
and corrects for that amount during its next ETA computation ({\tt fudge\_time()} in {\tt estimate.c}).
Landslide's data race analysis also includes a heuristic to avoid reporting ``too suspicious''
data-race candidates which it believes arise from the initialization pattern \cite{eraser}:
if a conflicting access pair is single-order (\cref{sec:quicksand-heuristics})
and also arose during a known synchronization API's {\tt init()} or {\tt destroy()} function,
Landslide will not message it to Quicksand,
at least not until it is reclassified as both-order.

To recognize the reallocation pattern
discussed in \cref{sec:quicksand-id-realloc}
during data race analysis,
Landslide includes a generation counter in its heap allocation tracking (\cref{sec:landslide-memory}).
Each heap allocation is given a unique ID,
and when evaluating whether two heap accesses can race,
the IDs of their containing blocks must match
({\tt was\_freed\_remalloced()} in {\tt memory.c}),
in addition to the other requirements of Happens-Before.
If the generations do not match,
Landslide sets the {\tt free\_re\_malloc} flag in the messaging protocol to Quicksand.
If the race is later observed in a reordering which avoids the reallocation pattern
(such as in \Cref{fig:recycle-bug}),
Landslide will report it as normal,
and Quicksand will promote it to a normal preemption point in the registry
({\tt pp\_new()} in {\tt pp.c}, ``{\tt for realsies}'' case).
Also included in this message is a flag to indicate whether a data race was found
nondeterministically (i.e., not on the first interleaving), such as described in \cref{sec:quicksand-pps}.

@@

\qrevision{In Quicksand, Iterative Deepening and data race analysis are intimately connected:
the former relies on the later to supply it with new preemption points, thereby refining its search for new concurrent behaviours,
while the latter relies on the former to thoroughly check all possible interleavings around its reported memory accesses and classify them as buggy or benign.
Despite this synergy, which is necessary for total verification soundness (\cref{sec:quicksand-soundness}),
each of these two techniques is a contribution in its own right when it comes to bug-finding performance.
Hence, this evaluation will measure not only the combined approach's full verification power,
but also the bug-finding performance of each technique separately,
as compared to state-of-the-art single-state-space ICB and DPOR.
I pose the following evaluation questions.}

\begin{enumerate}
	\item Does integrating data-race preemption points improve the accuracy of model checking?
		\begin{enumerate}
			\item Do data-race preemption points expose new bugs that couldn't be found with
				synchronization ones alone?
			\item Do data-race preemption points expose the same bugs
				the preempt-everywhere approach could find, faster?
			\item Does Quicksand provide more full verifications %of bug-free programs
				more quickly than the preempt-everywhere approach?
			% not related to new phrasing of Q1, and also the answer is "kinda sorta not really sometimes" anyway
			%\item Does Iterative Deepening find bugs faster in subset state spaces,
			%	even without data-race preemption points?
			\item How does the choice between Pure and Limited Happens-Before
				affect bug-finding and verification performance?
		\end{enumerate}
	\item Does testing alternate interleavings with model checking improve the accuracy of data race analysis?
		\begin{enumerate}
			\item Does Quicksand avoid false positives compared to single-execution Limited Happens-Before?
			\item Does Quicksand find data-race bugs that single-execution Pure Happens-Before or Limited Happens-Before alone would miss?
		\end{enumerate}
\end{enumerate}

@@

\begin{itemize}
	\item {\bf QS-Limited-HB}: Quicksand with Landslide configured to use Limited Happens-Before for its data race analysis ({\tt -H}),
	\item {\bf QS-Pure-HB}: Quicksand using Pure Happens-Before instead ({\tt -V}), and
	\item {\bf QS-Sync-Only}: Quicksand with initial preemption points only,
		as described in \cref{sec:quicksand-initial-pps},
		but never adding new ones from reported data races.
\end{itemize}

@@

\Cref{fig:dr-falsenegs}
%compares the types of data-race candidates necessary to expose each data-race bug in the test suite.
visualizes the difference between single-pass and model-checking-enabled data race analysis.
The first and third series represent the bugs found using preemption points from single-pass data-race candidates only,
% not entirely true, as portend could be given data-race traces from an MC
%but they don't do it in their paper, so i feel comfortable making this claim
i.e., the state-of-the-art approach used by RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}.
The second and fourth series show all data-race bugs Quicksand found,
which includes the former type as well as new bugs involving nondeterministic races.
QS-Limited-HB found a nice 69 data-race bugs in total, % nice
15 of which %could not be found with single-pass data-race candidates alone.
required nondeterministic data-race preemption points to expose.
QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race preemption points,
%requiring nondeterministic data-race preemption points
requiring them in 21 cases among its 44 total data-race bugs.
Moreover, although the frequency
of these nondeterministic races varies across the different test cases
(for example, almost all in {\tt broadcast\_test} were nondeterministic; almost none in {\tt mutex\_test}),
they are still at least present in all tests,
meaning it is not just an issue of writing ``better'' test cases to avoid them.

@@

In \cref{sec:quicksand-realloc} I showed the soundness of
suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
\Cref{tab:drstatistix}'s ``Realloc FPs'' column shows the total number of such data-race candidates for each test program,
totaling 965 across all tests.
% NB. Calculated this by grepping QS-LHB id-logs for "for realsies" (see quicksand pp.c).
Among these, only 64 were observed to avoid the reallocation in an alternate interleaving,
thereupon being promoted to real data-race preemption points.
\cref{sec:quicksand-realloc}'s proof guarantees the safety of pruning all state spaces resulting from the 901 others.
%
Among the 64 true data-races, %which initially fit the malloc-recycle pattern,
none exposed a new bug when used as a preemption point.
This suggests that for other data-race tools,
suppressing reallocation candidates may be a productive heuristic,
even if unsound without Iterative Deepening.
However, Quicksand was able to correctly identify the 64 violations of that heuristic (among 26 distinct tests),
and fall back to classifying them with DPOR.

@@

\subsubsection{Static data race analysis}

@@

\begin{itemize}
	\item Iterative Deepening (\cref{sec:quicksand-id}),
		an algorithm for model checkers to simultaneously test multiple state spaces,
		incorporating new preemption points identified with dynamic analysis on the fly.
	\item A proof of convergence (\cref{sec:quicksand-convergence}),
		showing that for any verification obtained under even the most extreme preemption strategy,
		Iterative Deepening with data race analysis provides an equivalently strong one,
		with far fewer preemption points necessary.
	\item A technique for suppressing certain false positive data-race candidates
		under Limited Happens-Before (\cref{sec:landslide-lhb})
		by identifying intervening {\tt malloc()} and {\tt free()} calls
		(\cref{sec:quicksand-id-realloc}),
		and a corresponding soundness proof when this technique is used under Iterative Deepening
		(\cref{sec:quicksand-realloc}).
	\item Quicksand (\cref{sec:quicksand-implementation}),
		an Iterative Deepening implementation which incorporates several heuristics for prioritizing
		which state spaces are most likely to uncover bugs % this claim is not scientifically justified
		or, should no bugs exist therein,
		which ones are most likely to complete within a user-specified fixed CPU budget,
		as informed by state space estimation (\cref{sec:landslide-estimate}).
	\item A 629-test evaluation of Quicksand
		against several prior state-of-the-art model checking approaches implemented in Landslide
		(\cref{sec:quicksand-eval}),
		which showed that Quicksand provides both faster bug-finding (\cref{sec:quicksand-eval-bugs})
		and more full verifications (\cref{sec:quicksand-eval-verif}),
		delivering ``the best of both worlds'' as promised,
		and also demonstrated the need for a bidirectional feedback loop
		between model checking and data race analysis
		(\cref{sec:quicksand-eval-nondets}).
\end{itemize}

>>>>

This chapter presents Quicksand,
an execution framework for model checking to manage this trade-off at run-time.
Given a fixed CPU budget,
representing the user's patience for testing,
Quicksand dynamically alters its preemption strategy based on data-race analysis
(\cite{tsan,fasttrack}, \cref{sec:landslide-datarace})
and optimizes the size of state spaces on the fly,
guided by state space estimation
(\cite{estimation}, \cref{sec:landslide-estimate}),
to best match that budget.
I will discuss the trade-off inherent in number of preemption points used
(\cref{sec:quicksand-motivation}),
introduce {\em Iterative Deepening},
the algorithm that Quicksand uses to automatically navigate that trade-off
(\cref{sec:quicksand-id}),
prove its soundness relative to the more expensive full verification approach
(\cref{sec:quicksand-soundness}),
and present a large evaluation of Quicksand against several
state-of-the-art approaches % implemented under Landslide
in which Quicksand performs best on both bug-finding and verification
(\cref{sec:quicksand-eval}).
% justifying the claim that model checkers should navigate said tradeoff at runtime
}

@@

\qrevisionminor{How should a model checker know to instrument this particular data race for preemption
in order to find the assertion failure lurking underneath?}
Committing in advance to
preempt on every instruction is certain to include this data race,
but invites massive state space explosion.
Even as DPOR helps to skip equivalent interleavings of non-conflicting transitions,
DPOR itself is $O(n^2)$ in the number of preemption points in a single execution,
which is not compatible with such an approach.
Accordingly, stateless model checkers must find more efficient ways to be able to uncover bugs such as these.
%
In related work,
Portend~\cite{portend} proposed to combine data-race analysis with preemption-driven artificial scheduling,
although it obtains its data-race candidates from a stand-alone, single-pass analysis.
In order to
%make sure we
identify every
%one of a program's
data race that could possibly arise under the given test case,
a model checker must check many different interleavings to begin with,
perform the Portend approach for every data race it finds,
which may in turn uncover more data races
(hidden in flow control paths reachable only through interleavings of the first race, perhaps),
and then continue model checking those multiple races together
in a
%sort of
bidirectional feedback loop between the two algorithms.
Upcoming,
\cref{sec:quicksand-id} will show how I achieve this in Quicksand, and
\cref{sec:quicksand-soundness} will justify the technique's formal verification power.
In the evaluation, %(\cref{sec:quicksand-eval}),
\qrevisionminor{\cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}
will show that Quicksand strikes a healthy balance between fast bug-finding and full verification,
and
\cref{sec:quicksand-eval-nondets}
%in particular
will justify the need
for such a feedback loop %between model checking and data-race analysis.
by showing that many
data races require model checking to reliably detect.
%bugs require nondeterministic data races to expose.
}

@@

For the remainder of the chapter, I will use Iterative Deepening to refer to the algorithm in the abstract,
which could in principle apply to any stateless model checking domain,
and Quicksand to refer to the specific implementation,
which relies on data-race analysis and specific heuristics to optimize its testing approach
for kernels and thread libraries.
I will also henceforth refer to each unique set of preemption points as a {\em job}.

@@

%Note that
Iterative Deepening is a wrapper algorithm around stateless model checking.
A model checker is still used to test each state space, and other reduction techniques such as DPOR
are still applicable in each.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
This chapter
will focus on run-time data-race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
The next section (\cref{sec:quicksand-soundness})
will prove that in fact,
in addition to statically-known synchronization preemption points,
this suffices to provide at least as strong verification guarantees as any other possible preemption point set.

@@

However, in many tests of even modestly-sized programs,
full verification is not feasible,
and focusing on the maximal state space alone is likely to be fruitless.
%
Hence, Iterative Deepening also allows for prioritizing subset jobs
based on number of preemption points, ETA, and whether data-race candidates are included among their preemption points.
It relies on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing a large fraction of interleavings for each.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only her total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.
Quicksand seeks to maximize completed state spaces,
as each one serves as a guarantee that all possible interleavings therein were tested;
\cref{sec:quicksand-discussion-partial} discusses some limitations of this approach.
The next three subsections will show how to schedule these smaller jobs
based on their preemption points and ETAs.

@@

For kernel-level testing, interrupt-disabling is analogous to locking,
so preemptions must also be introduced
just before a disable-interrupt opcode (on x86, {\tt cli})
and just after interrupts are re-enabled (on x86, {\tt sti}).
During data-race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
(note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
\footnote{Some kernels disable preemption without disabling interrupts,
which can be communicated to the model checker using manual annotations,
and must be treated similarly.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}
Quicksand is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, it doesn't test every intermediate subset such as $\{lock,sti\}$,
which would result in $2^p$ jobs right off the bat,
although this could potentially be improved in future work (\cref{sec:quicksand-discussion-subsets}).

@@

As discussed in \cref{sec:quicksand-pps},
%runtime data-race detection (\cref{sec:background-datarace})
%may find candidate unsynchronized memory conflicts that should be investigated further.
data races may beget new interleavings not reachable by preempting on synchronization API boundaries alone.
Because each data race indicates an access pair that can interleave at instruction granularity,
different program behaviour may arise if the threads are preempted just before the racing instructions,
some of which the programmer may not have even expected, i.e., be bugs,
and it is logical to apply model checking to find or verify the absence of such bugs.
%it is logical to re-execute the test and issue preemptions just before those instructions
%to test alternate thread interleavings.

@@

\begin{figure}[t]
	\begin{center}
        \includegraphics[width=0.9\textwidth]{dr-jobs-v2.pdf}
	\end{center}
	\caption[Quicksand incorporates data races as new preemption points at run-time.]
		{Quicksand incorporates data races as new preemption points at run-time
		by managing the exploration of multiple state spaces,
		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
                %When an access pair is reported as a data-race candidate,
                When a data race is reported,
                a new preemption point is added for each involved memory access,
                and new jobs are added for later testing,
                corresponding to different combinations thereof with the existing preemption points.}
        \label{fig:new-dr-jobs}
\end{figure}

@@

% TODO: fix syntax hilight
% TODO check figure placement
% TODO check figure camption
\begin{figure}[t]
	\begin{center}
	\begin{tabular}{rll}
		& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
		& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
		\\
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		2 & \texttt{\hilight{olivegreen}{free}(x);} \\
		3 & & \texttt{\hilight{commentblue}{// x's memory reallocated}} \\
		4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
		6 & & \texttt{publish(y);} \\
		7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
	\end{tabular}
	\end{center}
	% TODO: short ver
	\caption{A common execution pattern with {\tt malloc()} that produces false positive data-race candidates.}
	\label{fig:recycle}
\end{figure}

@@

The ability to disregard reallocation false positives is unique to Iterative Deepening.
When limited to a single test execution, suppressing any data-race candidate matching this pattern is unsound.
Consider the more unusual program in Figure~\ref{fig:recycle-bug},
in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
Here, reordering lines~6-7 before line~3 will allow {\tt x} and {\tt x2} to race.
Discarding the data race report as a false positive after checking just this one execution
would overlook such a bug,
but Iterative Deepening is guaranteed to explore the alternate interleaving,
in which the true data race will show up without {\tt free()} and {\tt malloc()} interposing,
so it is safe to suppress at first, as I will prove in \cref{sec:quicksand-realloc}.
Moreover,
in the context of Iterative Deepening, being able to discard certain data-race candidates
allows Quicksand to skip exploring some entire state spaces,
and hence run fewer Landslides overall;
this is analogous to DPOR's ability to skip equivalent interleavings within a single Landslide instance.
Upcoming in the evaluation, \cref{sec:quicksand-eval-verif}'s \Cref{tab:drstatistix}
will show how many redundant state spaces Quicksand is able to prune with this technique.

@@

% TODO: check spcaing between defs/lemmas/etc and surrounding text
\begin{definition}[Reachability]
	A data-race candidate, and its associated preemption point(s),
	are reachable if it will be identified by a model checker
	configured to preempt only on already-reachable preemption points.
\end{definition}

@@

If the verbose option ({\tt -v}) is supplied,
Quicksand will also print one line per interleaving tested by all its Landslide instances,
showing the number of branches tested so far, the estimated percent progress, and the ETA,
as shown earlier in \cref{sec:landslide-estimate}.
This produces a lot more output, and can make progress reports hard to read as they scroll off the screen quickly,
but the author personally finds this mode less disorienting than ten seconds of pure silence between each progress report.
Of course, future work could improve this with a GUI, or at least a split-screen console view.
It will also cause the progress reports to report more detailed information,
such as which preemption points are nondeterministic data races (\cref{sec:quicksand-pps})
and number of reallocation false positives suppressed (\cref{sec:quicksand-id-realloc}).

@@

The other end of the messaging protocol (\cref{sec:quicksand-impl-mc})
is implemented in {\tt messaging.c}.
When Quicksand suspends Landslide,
it detects how much time it spent asleep,
and corrects for that amount during its next ETA computation ({\tt fudge\_time()} in {\tt estimate.c}).
Landslide's data-race analysis also includes a heuristic to avoid reporting ``too suspicious''
data-race candidates which it believes arise from the initialization pattern \cite{eraser}:
if a conflicting access pair is single-order (\cref{sec:quicksand-heuristics})
and also arose during a known synchronization API's {\tt init()} or {\tt destroy()} function,
Landslide will not message it to Quicksand,
at least not until it is reclassified as both-order.

To recognize the reallocation pattern
discussed in \cref{sec:quicksand-id-realloc}
during data-race analysis,
Landslide includes a generation counter in its heap allocation tracking (\cref{sec:landslide-memory}).
Each heap allocation is given a unique ID,
and when evaluating whether two heap accesses can race,
the IDs of their containing blocks must match
({\tt was\_freed\_remalloced()} in {\tt memory.c}),
in addition to the other requirements of Happens-Before.
If the generations do not match,
Landslide sets the {\tt free\_re\_malloc} flag in the messaging protocol to Quicksand.
If the race is later observed in a reordering which avoids the reallocation pattern
(such as in \Cref{fig:recycle-bug}),
Landslide will report it as normal,
and Quicksand will promote it to a normal preemption point in the registry
({\tt pp\_new()} in {\tt pp.c}, ``{\tt for realsies}'' case).
Also included in this message is a flag to indicate whether a data race was found
nondeterministically (i.e., not on the first interleaving), such as described in \cref{sec:quicksand-pps}.

@@

\qrevision{In Quicksand, Iterative Deepening and data-race analysis are intimately connected:
the former relies on the later to supply it with new preemption points, thereby refining its search for new concurrent behaviours,
while the latter relies on the former to thoroughly check all possible interleavings around its reported memory accesses and classify them as buggy or benign.
Despite this synergy, which is necessary for total verification soundness (\cref{sec:quicksand-soundness}),
each of these two techniques is a contribution in its own right when it comes to bug-finding performance.
Hence, this evaluation will measure not only the combined approach's full verification power,
but also the bug-finding performance of each technique separately,
as compared to state-of-the-art single-state-space ICB and DPOR.
I pose the following evaluation questions.}

\begin{enumerate}
	\item Does integrating data-race preemption points improve the accuracy of model checking?
		\begin{enumerate}
			\item Do data-race preemption points expose new bugs that couldn't be found with
				synchronization ones alone?
			\item Do data-race preemption points expose the same bugs
				the preempt-everywhere approach could find, faster?
			\item Does Quicksand provide more full verifications %of bug-free programs
				more quickly than the preempt-everywhere approach?
			% not related to new phrasing of Q1, and also the answer is "kinda sorta not really sometimes" anyway
			%\item Does Iterative Deepening find bugs faster in subset state spaces,
			%	even without data-race preemption points?
			\item How does the choice between Pure and Limited Happens-Before
				affect bug-finding and verification performance?
		\end{enumerate}
	\item Does testing alternate interleavings with model checking improve the accuracy of data-race analysis?
		\begin{enumerate}
			\item Does Quicksand avoid false positives compared to single-execution Limited Happens-Before?
			\item Does Quicksand find data-race bugs that single-execution Pure Happens-Before or Limited Happens-Before alone would miss?
		\end{enumerate}
\end{enumerate}

@@

\begin{itemize}
	\item {\bf QS-Limited-HB}: Quicksand with Landslide configured to use Limited Happens-Before for its data-race analysis ({\tt -H}),
	\item {\bf QS-Pure-HB}: Quicksand using Pure Happens-Before instead ({\tt -V}), and
	\item {\bf QS-Sync-Only}: Quicksand with initial preemption points only,
		as described in \cref{sec:quicksand-initial-pps},
		but never adding new ones from reported data races.
\end{itemize}

@@

\Cref{fig:dr-falsenegs}
%compares the types of data-race candidates necessary to expose each data-race bug in the test suite.
visualizes the difference between single-pass and model-checking-enabled data-race analysis.
The first and third series represent the bugs found using preemption points from single-pass data-race candidates only,
% not entirely true, as portend could be given data-race traces from an MC
%but they don't do it in their paper, so i feel comfortable making this claim
i.e., the state-of-the-art approach used by RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}.
The second and fourth series show all data-race bugs Quicksand found,
which includes the former type as well as new bugs involving nondeterministic races.
QS-Limited-HB found a nice 69 data-race bugs in total, % nice
15 of which %could not be found with single-pass data-race candidates alone.
required nondeterministic data-race preemption points to expose.
QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race preemption points,
%requiring nondeterministic data-race preemption points
requiring them in 21 cases among its 44 total data-race bugs.
Moreover, although the frequency
of these nondeterministic races varies across the different test cases
(for example, almost all in {\tt broadcast\_test} were nondeterministic; almost none in {\tt mutex\_test}),
they are still at least present in all tests,
meaning it is not just an issue of writing ``better'' test cases to avoid them.

@@

In \cref{sec:quicksand-realloc} I showed the soundness of
suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
\Cref{tab:drstatistix}'s ``Realloc FPs'' column shows the total number of such data-race candidates for each test program,
totaling 965 across all tests.
% NB. Calculated this by grepping QS-LHB id-logs for "for realsies" (see quicksand pp.c).
Among these, only 64 were observed to avoid the reallocation in an alternate interleaving,
thereupon being promoted to real data-race preemption points.
\cref{sec:quicksand-realloc}'s proof guarantees the safety of pruning all state spaces resulting from the 901 others.
%
Among the 64 true data races, %which initially fit the malloc-recycle pattern,
none exposed a new bug when used as a preemption point.
This suggests that for other data-race tools,
suppressing reallocation candidates may be a productive heuristic,
even if unsound without Iterative Deepening.
However, Quicksand was able to correctly identify the 64 violations of that heuristic (among 26 distinct tests),
and fall back to classifying them with DPOR.

@@

\subsubsection{Static data-race analysis}

@@

\begin{itemize}
	\item Iterative Deepening (\cref{sec:quicksand-id}),
		an algorithm for model checkers to simultaneously test multiple state spaces,
		incorporating new preemption points identified with dynamic analysis on the fly.
	\item A proof of convergence (\cref{sec:quicksand-convergence}),
		showing that for any verification obtained under even the most extreme preemption strategy,
		Iterative Deepening with data-race analysis provides an equivalently strong one,
		with far fewer preemption points necessary.
	\item A technique for suppressing certain false positive data-race candidates
		under Limited Happens-Before (\cref{sec:landslide-lhb})
		by identifying intervening {\tt malloc()} and {\tt free()} calls
		(\cref{sec:quicksand-id-realloc}),
		and a corresponding soundness proof when this technique is used under Iterative Deepening
		(\cref{sec:quicksand-realloc}).
	\item Quicksand (\cref{sec:quicksand-implementation}),
		an Iterative Deepening implementation which incorporates several heuristics for prioritizing
		which state spaces are most likely to uncover bugs % this claim is not scientifically justified
		or, should no bugs exist therein,
		which ones are most likely to complete within a user-specified fixed CPU budget,
		as informed by state space estimation (\cref{sec:landslide-estimate}).
	\item A 629-test evaluation of Quicksand
		against several prior state-of-the-art model checking approaches implemented in Landslide
		(\cref{sec:quicksand-eval}),
		which showed that Quicksand provides both faster bug-finding (\cref{sec:quicksand-eval-bugs})
		and more full verifications (\cref{sec:quicksand-eval-verif}),
		delivering ``the best of both worlds'' as promised,
		and also demonstrated the need for a bidirectional feedback loop
		between model checking and data-race analysis
		(\cref{sec:quicksand-eval-nondets}).
\end{itemize}
